---
title: 'ESM 244 HW 3 Task 2: Hierarchical Clustering of Watershed Sites'
author: "Sarah Hamilton"
date: "2/18/2022"
output:
  html_document:
    code_folding: hide
    theme: yeti
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: yes
    smooth_scroll: yes
    number_sections: yes
editor_options:
  chunk_output_type: console
---

# Introduction

3-4 sentence intro

Data Citation: 

# Hierarchical Cluster Analysis on Water Chemistry

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(palmerpenguins)
library(here)

library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
```

```{r}
# Read in and wrangle the data 

water <- read_csv(here("data", "sbc_lter_registered_stream_chemistry.csv")) %>%
  na_if(-999) %>%
  drop_na %>%
  select(-timestamp_local)

# Group the data by site and get the mean value for each site
water_sites <- water %>%
  group_by(site_code) %>%
  summarize(
    nh4_uM_mean = mean(nh4_uM),
    no3_uM_mean = mean(no3_uM),
    po4_uM_mean = mean(po4_uM),
    tdn_uM_mean = mean(tdn_uM),
    tdp_uM_mean = mean(tdp_uM),
    tpc_uM_mean = mean(tpc_uM),
    tpn_uM_mean = mean(tpn_uM),
    tpp_uM_mean = mean(tpp_uM),
    tss_mgperLiter_mean = mean(tss_mgperLiter),
    spec_cond_uSpercm_mean = mean(spec_cond_uSpercm)
  )

# Scale the data
water_sites_scale <- water_sites %>%
  select(-site_code) %>%
  scale()

# Rename the rows to be the site name
rownames(water_sites_scale) <- water_sites$site_code
```

## Complete Linkage Dendrogram 

```{r}
# Calculate distance matrix
water_dist <- dist(water_sites_scale, method = 'euclidean')

# Complete linkage hierarchical clustering
water_hc_complete <- hclust(water_dist, method = 'complete')

# Plot a dendrogram
ggdendrogram(water_hc_complete, rotate = TRUE) +
  theme_minimal() +
  labs(x = "Site", y = "Distance")
```

**Figure 1.** Dendrogram for complete linkage hierarchical clustering of watershed sites based on stream water chemistry.

## Single Linkage Dendrogram

```{r}
# Single linkage hierarchical clustering
water_hc_single <- hclust(water_dist, method = 'single')

# Plot a dendrogram
ggdendrogram(water_hc_single, rotate = TRUE) +
  theme_minimal() +
  labs(x = "Site", y = "Distance")
```

**Figure 2.** Dendrogram for single linkage hierarchical clustering of watershed sites based on stream water chemistry."

## Tanglegram

```{r}
# Plot a tanglegram
dend_complete <- as.dendrogram(water_hc_complete)
dend_single <- as.dendrogram(water_hc_single)

tanglegram(dend_complete, dend_single)
```

**Figure 3.** Tanglegram for complete and single linkage hierarchical clustering of watershed sites based on stream water chemistry. 

# Summary

The hierarchical clustering tells us the following:

- asdf

- asdf

- asdf

TO BE DELETED: A brief summary (nicely formatted and professionally written bulletpoints are fine) highlighting some major takeaways from your hierarchical clustering analysis that can be gleaned from the dendrogram(s).
